A saturation model for impaired learning with enhanced plasticity: Supplementary material

### 2013-02-25 Notes/MHCsupp.tex
MHC supplemental

% -*- TeX -*- -*- UK -*-
% ----------------------------------------------------------------
% arXiv Paper ************************************************
%
% Subhaneil Lahiri's template
%
% Before submitting:
%    Comment out hyperref
%    Comment out showkeys
%    Replace \input{mydefs.tex} with its contents
%       or include mydefs.tex in zip/tar file
%    Replace \input{newsymb.tex} with its contents
%       or include newsymb.tex in zip/tar file
%    Put this file, the .bbl file, any picture or
%       other additional files and natbib.sty
%       file in a zip/tar file
%
% **** -----------------------------------------------------------
\documentclass[12pt]{article}
%Preamble:
\usepackage{a4wide}
\input{sl_preamble.tex}
\input{sl_graphics_preamble.tex}
\graphicspath{{"Figures/"}}
%
\usepackage{array}
\newcounter{tablerow}
\newcommand{\tablerowautorefname}{row}
\newenvironment{tabularn}[1]{\setcounter{tablerow}{0}\begin{tabular}{#1>{\refstepcounter{tablerow}}l}}{\end{tabular}}
%\usepackage{etoolbox}
%\AtBeginEnvironment{tabular}{\setcounter{tablerow}{0}}
%
% >> Only for drafts! <<
\usepackage[notref,notcite]{showkeys}
% ----------------------------------------------------------------
%\numberwithin{equation}{section}
%\renewcommand{\baselinestretch}{1.5}
% ----------------------------------------------------------------
% New commands etc.
\input{sl_definitions.tex}
\input{sl_symbols.tex}
%matrices
\newcommand{\inv}{^{-1}}
\newcommand{\dg}{^\mathrm{dg}}
\newcommand{\trans}{^\mathrm{T}}
\newcommand{\I}{\mathbf{I}}
%vec of ones
\newcommand{\onev}{\mathbf{e}}
%mat of ones
\newcommand{\onem}{\mathbf{E}}
%Markov matrix
\newcommand{\MM}{\mathbf{Q}}
%equilibrium distribution
\newcommand{\pr}{\mathbf{p}}
\newcommand{\eq}{\pr^\infty}
%first passage times
\newcommand{\fpt}{\mathbf{T}}
%off-diag first passage times
\newcommand{\fptb}{\overline{\fpt}}
%fundamental matrix
\newcommand{\fund}{\mathbf{Z}}
%other symbols
\newcommand{\Pb}{\mathbf{P}}
\newcommand{\D}{\mathbf{D}}
\newcommand{\pib}{\boldsymbol{\pi}}
\newcommand{\Lb}{\boldsymbol{\Lambda}}
\newcommand{\w}{\mathbf{w}}
\newcommand{\W}{\mathbf{W}}
\newcommand{\frg}{\W^\mathrm{F}}
\newcommand{\M}{\mathbf{M}}
\newcommand{\F}{\boldsymbol{\Phi}}
\DeclareMathOperator{\SNR}{SNR}
\newcommand{\CS}{\mathcal{S}}
\newcommand{\CA}{\mathcal{A}}
\newcommand{\CB}{\mathcal{B}}
\newcommand{\comp}{^\mathrm{c}}
\newcommand{\pot}{^{\text{pot}}}
\newcommand{\dep}{^{\text{dep}}}
\newcommand{\potdep}{^{\text{pot/dep}}}
\newcommand{\norm}{_0}
\newcommand{\inc}{_{\text{inc}}}
\newcommand{\dec}{_{\text{dec}}}
\newcommand{\incdec}{_{\text{inc/dec}}}
\newcommand{\wt}{_{\text{WT}}}
\newcommand{\ko}{_{\text{D$^\mathrm{b}$K$^\mathrm{b}$-/-}}}
\newcommand{\KO}{D$^\mathrm{b}$K$^\mathrm{b}$-/-}
\newcommand{\tpre}{t_{\text{pre}}}
\newcommand{\ttrain}{t_{\text{train}}}
\newcommand{\lmax}{_{\text{max}}}
\newcommand{\lmin}{_{\text{min}}}
% ----------------------------------------------------------------
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Title info:
\title{Models of VOR learning in MHC knockout mice}
%
% Author List:
%
\author{Subhaneil Lahiri
%
}

@@

\begin{abstract}
  We see if we can model VOR gain increase and decrease learning in mice with a knockout in MHC as well as wild-type.
\end{abstract}

@@

\section{Introduction and summary}\label{sec:intro}

We will see if we can capture the essential features of VOR gain-increase learning in wild type and MHC \KO\ mutants and wild-type, with and without gain-decrease pre-training.
We will try to do this using models of complex synapses, such as the cascade model \cite{Fusi2005cascade} or the multistate model \cite{amit1994learning,Fusi2007multistate}.\footnote{The original definition of the multistate model had a synaptic weight that varied linearly along the chain. We will refer to this as the linear multistate model. We will also look at the step-multistate model, for which the synaptic weight to take two values.}
We will also construct a new model in \autoref{sec:pooledmodel} that incorporates the idea that biasing the synapses in one direction could deplete some resource, thereby impairing further plasticity in the same direction in nearby synapses.

These models all assume independent, statistically identical synapses.
This has the benefit of freeing us from modelling the neurons and their networks, allowing us to focus on the synapses themselves.
However, we will lose certain aspects of real synapses.
For example, during gain-increase learning, certain synapses will be potentiated more often and others will be depressed more often.
In these models, however, synaptic identity is not kept track of, so training can only be described by changing the overall levels of potentiation and depression.

As the MHC \KO\ mutation lowers the threshold for LTD, we will model these mice by increasing the probabilities for internal synaptic transitions associated with depression, whilst leaving potentiation unchanged.

We will be parsimonious in our modelling.
We will assume that the wild-type has a symmetry between potentiation and depression and we will treat gain-increase and decrease training in a symmetric way.
We will also assume that the relative rates of potentiation/depression in the mutant and wild-type are the same and that the overall rate of plasticity events is the same in all situations.
None of these assumptions are necessary or justified experimentally (although \cite{Fusi2007multistate} notes that imbalance between potentiation and depression can drastically impair memory).
They do reduce the number of adjustable parameters we have, simplifying searches for good models.
Presumably, relaxing these assumptions and enlarging the parameter space would allow for a better fit to data, but the number of parameters would start to look large compared to the number of features we're trying to match.

If we imagine something like an LN model for the Purkinje cell, the output would depend on some linear combination of the synaptic weights (weighted by the activities of the corresponding parallel fibres).
If we're not keeping track of synaptic identity, the most natural linear combination to use would be an equal sum of them all.
Thus we will quantify learning by the change in the mean synaptic weight.

The output of the Purkinje cell, and the behavioural output of the animal, will depend on this quantity in some complicated nonlinear way.
Modelling this is far beyond the scope of this work.
Instead we will look at more qualitative features, asking whether learning is slower or faster in some situation compared to another.
Provided that the nonlinearity is monotonic, these types of questions will not be affected by it.

Within the parameter range explored, we will see that synaptic complexity is essential.
The simple two-state model always shows enhanced learning in the mutant (see \autoref{fig:binary}), in contradiction with experiment.
It seems that the enhanced depressing transition more than compensates for saturation effects.

A similar thing happens for the linear multistate model (see \autoref{fig:multistate_lin}), but the wild-type catches up and overtakes the mutant very quickly.

On the other hand, the cascade, step-multistate and pooled resource models all have parameter ranges where the mutant has impaired learning without pre-training (see \autoref{fig:cascade_short}, \autoref{fig:cascade_long}, \autoref{fig:multistate_weak}, \autoref{fig:multistate_strong} and \autoref{fig:pooled_scarce}).
We have found parameter ranges where the reverse happens for the multistate (see discussion below \eqref{eq:multiKOnopre}) and the pooled resource model (see \autoref{fig:pooled_plenty}).
These rely on properties of the mutation, so it would be difficult to explore with further experiments.

All of these models show pre-training enhancing learning for the mutant, as expected.
It is harder to find regimes where pre-training impairs learning for the wild-type.
In the cascade model, this required a much longer duration for pe-training (see \autoref{fig:cascade_long}).
In the step-multistate model, this required stronger training, which resulted in gain-decrease learning being faster for the mutant then wild-type (see \autoref{fig:multistate_strong}), unlike what is seen experimentally.
It is possible that these features could be improved by lessening the symmetry between gain-increase and decrease or potentiation and depression, or by allowing the rate of plasticity events to vary.
For the linear multistate and pooled resource models, we haven't found any parameter regime where pre-training impairs learning for the wild-type, and we discuss why it seems unlikely in \autoref{sec:pooled}.

There seems to be two take home messages.
First, impaired learning without pre-training in the mutant requires synaptic complexity, as it is not seen in the simple two-state model.
Second, impaired learning with pre-training in the wild-type requires a form of metaplasticity where repeated potentiation makes subsequent depression less likely.
This is a feature of the cascade model and our version of the multistate model, but not the pooled resource model or linear multistate.

\section{The setup}\label{sec:setup}

@@

We make the following assumptions:
\begin{itemize}
  \item There are $N$ identical synapses with $M$ internal functional states.
  \item The states of different synapses are independent of each other.
  \item The synapses that are eligible for plasticity are chosen randomly.
  \item The potentiating/depressing plasticity event timings are distributed as Poisson processes with rates $rf\potdep$, where $f\pot+f\dep=1$.
      In other words, $r$ is the total rate of plasticity events per synapse and $f\potdep$ are the fraction of these events that are potentiating/depressing.
  \item Potentiation and depression are described by Markov processes with transition probabilities $\M\potdep$.
  \item The synaptic weights of the internal states are given by the column vector $\w$. This can only take two values that we can call $\pm1$, except for the pooled resource model that we will discuss in \autoref{sec:pooledmodel} and \autoref{sec:pooled}.
\end{itemize}

@@

With only two possible synaptic weights, the distribution of synaptic weights is completely described by the mean, $\pr(t)\w$.

We will describe the effect of training by the decrease in mean synaptic weight:
%
\begin{equation}\label{eq:learning}
  L(t) = \prn{\pr(0)-\pr(t)}\w.
\end{equation}
%
The behavioural output (VOR gain) will be some non-linear function of the synaptic weights, so the best we can hope for is to reproduce qualitative features of the experiment.

The MHC mutant has a lower threshold for depression.
We can model this by changing $\W\dep\wt$ to $\W\dep\ko$, which should have larger matrix elements.

%QUESTION: Should $f\pot\wt=f\pot\ko$?
%
%If they are equal, this would change the mean synaptic weight in equilibrium.
%This seems like it could affect the ability of the network to perform its function, whereas, in reality, the wild-type and mutant have the same baseline performance.
%In any case, if the synaptic weights are different, the electrical activity will be different, and there will be no reason to expect the same statistics for potentiation or depression.
%In setting $f\pot\wt=f\pot\ko$, we are assuming that the rest of the brain compensates for the difference in synaptic weights somehow.
%
%On the other hand, one could imagine adjusting $f\pot\wt$ and $f\pot\ko$ so that $\eq\wt\w = \eq\ko\w$.
%But, now that the synaptic weights are the same, the electrical activity will be the same, and there will be no reason to expect different statistics for potentiation or depression.
%If we did this for the multistate or two-state models, we'd remove all differences between the wild-type and \KO\ mutant, so we won't.

We are also assuming that the relation between VOR gain and mean synaptic weight is the same for mutant ant wild-type, except for a linear offset to compensate for the difference in equilibrium weights mentioned above.

\section{Simulations}\label{sec:sims}

\subsection{Models and parameters}

We will look at three different models, the cascade model (see \cite{Fusi2005cascade} and \autoref{fig:models}\ref{fig:cascade_model}), the multistate model (see \cite{amit1994learning,Fusi2007multistate} and \autoref{fig:models}\ref{fig:multistate_model}) and the two-state model (which can be thought of as a special case of the previous two, see \autoref{fig:models}\ref{fig:binary_model}).
We will also look at a new, pooled resource model that we will define below in \autoref{sec:pooledmodel}.

@@

In each case, we set $f\pot\norm=\half$, $f\pot\inc=f\pot\norm+\Delta f$ and $f\pot\dec=f\pot\norm-\Delta f$.
We use the same values for wild-type and mutant and treat gain-increase and decrease symmetrically.
The values of all these parameters are listed in \autoref{tab:params}.

\begin{figure}
 \begin{center}
 \begin{myenuma}
  \item\aligntop{\includegraphics[height=2cm]{cascade.svg}}\label{fig:cascade_model}\hspace{0.5cm}
  \item\aligntop{\includegraphics[height=1.7cm]{multistate.svg}}\label{fig:multistate_model}\hspace{0.5cm}
  \item\aligntop{\includegraphics[height=1.7cm]{binary.svg}}\label{fig:binary_model}
 \end{myenuma}
 \end{center}
  \caption[Transition probabilities for different models]{Transition probabilities for different models.
  (\ref{fig:cascade_model}) In the cascade model, the transition probabilities decay geometrically with a parameter $x$ (see \cite{Fusi2005cascade}).
  (\ref{fig:multistate_model}) In the multistate model the transition probabilities for potentiation/depression are all equal and it is parameterised by these two values.
  (\ref{fig:binary_model}) The two-state model is parameterised by the two transition probabilities.} \label{fig:models}
\end{figure}

@@

This model is parameterised by the range of values, $q\in[q\lmin,q\lmax]$, for potentiation and depression.
We will use the same values for potentiation and depression in the wild-type as well as potentiation in the \KO\ mutant.
We will use larger values for depression in the \KO\ mutant.

\subsection{Results}\label{sec:results}

@@

@@

\subsubsection{Cascade model}\label{sec:cascade}

\subsubsection{Step-multistate model}\label{sec:multistate}

@@

\subsubsection{Linear multistate model}\label{sec:multistate_lin}

In this section, we will consider the multistate model as originally defined in \cite{amit1994learning}, \ie with linearly varying synaptic weight:
%
\begin{equation}\label{eq:multistateLinWeight}
  \w_i = \frac{2i-M-1}{M-1}.
\end{equation}
%
The numerical results can be seen in \autoref{fig:multistate_lin}, but we can get some analytic insight into this model as well.

\subsubsection{Two-state model}\label{sec:binary}

@@

\subsubsection{Pooled resource model}\label{sec:pooled}

@@

%\subsection*{Acknowledgements}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection*{Appendices}
%\appendix
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\documentclass[12pt]{article}
%Preamble:
\usepackage{a4wide}
\input{sl_preamble.tex}
\input{sl_graphics_preamble.tex}
\graphicspath{{"Figures/"}}
%
\usepackage{array}
\newcounter{tablerow}
\newcommand{\tablerowautorefname}{row}
\newenvironment{tabularn}[1]{\setcounter{tablerow}{0}\begin{tabular}{#1>{\refstepcounter{tablerow}}l}}{\end{tabular}}
%\usepackage{etoolbox}
%\AtBeginEnvironment{tabular}{\setcounter{tablerow}{0}}
%
% ----------------------------------------------------------------
% New commands etc.
\input{sl_definitions.tex}
\input{sl_symbols.tex}
%matrices
\newcommand{\I}{\mathbf{I}}
%vec of ones
%equilibrium distribution
\newcommand{\pr}{\mathbf{p}}
\newcommand{\eq}{\pr^\infty}
%other symbols
\newcommand{\w}{\mathbf{w}}
\newcommand{\W}{\mathbf{W}}
\newcommand{\frg}{\W^\mathrm{F}}
\newcommand{\M}{\mathbf{M}}
\newcommand{\F}{\boldsymbol{\Phi}}
\newcommand{\pot}{^{\text{pot}}}
\newcommand{\dep}{^{\text{dep}}}
\newcommand{\potdep}{^{\text{pot/dep}}}
\newcommand{\norm}{_0}
\newcommand{\inc}{_{\text{inc}}}
\newcommand{\dec}{_{\text{dec}}}
\newcommand{\incdec}{_{\text{inc/dec}}}
\newcommand{\wt}{_{\text{WT}}}
\newcommand{\ko}{_{\text{D$^\mathrm{b}$K$^\mathrm{b}$-/-}}}
\newcommand{\KO}{D$^\mathrm{b}$K$^\mathrm{b}$-/-}
\newcommand{\tpre}{t_{\text{pre}}}
\newcommand{\ttrain}{t_{\text{train}}}
\newcommand{\lmax}{_{\text{max}}}
\newcommand{\lmin}{_{\text{min}}}
% ----------------------------------------------------------------
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Title info:
\title{A saturation model for impaired learning with enhanced plasticity: Supplementary material}
%
%% Author List:
%%
%\author{An Author$^a$, Another Author$^b$ and Yet Another Author$^c$\\
%%
%\small{\emph{$^a$ Address 1}}\\
%\small{\emph{$^b$ Address 2}}\\
%\small{\emph{$^c$ Address 3}}\\
%}

@@

%\begin{abstract}
%  We see if we can model VOR gain increase and decrease learning in mice with a knockout in MHC as well as wild-type.
%\end{abstract}

@@

\section{Mathematical formalism}\label{sec:setup}

@@

We make the following assumptions:
\begin{itemize}
  \item There are $N$ identical synapses with $M$ internal functional states.
  \item The states of different synapses are independent of each other.
  \item The synapses that are eligible for plasticity are chosen randomly.
  \item The potentiating/depressing plasticity event timings are distributed as Poisson processes with rates $rf\potdep$, where $f\pot+f\dep=1$.
      In other words, $r$ is the total rate of plasticity events per synapse and $f\potdep$ are the fraction of these events that are potentiating/depressing.
  \item Potentiation and depression are described by Markov processes with transition probabilities $\M\potdep$.
  \item The synaptic weights of the internal states are given by the column vector $\w$. This can only take values in a finite range that we can shift to $\pm1$. Most of the models will only use the two extreme values.
\end{itemize}

@@

For models with only two possible synaptic weights, the distribution of synaptic weights is completely described by the mean, $\pr(t)\w$.

The MHC mutant has a lower threshold for depression.
We can model this by changing $\M\dep\wt$ to $\M\dep\ko$, which should have larger off-diagonal matrix elements.

We will look at several different models,
the multistate model\footnote{The original definition of the multistate model had a synaptic weight that varied linearly along the chain. We will refer to this as the linear multistate model. We will also look at the step-multistate model, for which the synaptic weight takes only two values.} (see \cite{amit1994learning,Fusi2007multistate} and \autoref{fig:models}\ref{fig:multistate_model}),
and the two-state model (which can be thought of as a special case of the multistate model, see \autoref{fig:models}\ref{fig:binary_model}),
and the cascade model (see \cite{Fusi2005cascade} and \autoref{fig:models}\ref{fig:cascade_model}).
We will also look at a new, pooled resource model that we will define below in \autoref{sec:pooledmodel}.

@@

The values of all these parameters are listed in \autoref{tab:params}.

\begin{figure}
 \begin{center}
 \begin{myenuma}
  \item\aligntop{\includegraphics[height=1.7cm]{multistate.svg}}\label{fig:multistate_model}\hspace{0.5cm}
  \item\aligntop{\includegraphics[height=1.7cm]{binary.svg}}\label{fig:binary_model}\hspace{0.5cm}
  \item\aligntop{\includegraphics[height=2cm]{cascade.svg}}\label{fig:cascade_model}
 \end{myenuma}
 \end{center}
  \caption[Transition probabilities for different models]{Transition probabilities for different models.
  Potentiation induces transitions indicated by solid red arrows, depression indicated by dashed blue arrows.
  States of strong/weak synaptic weight indicated by red/blue circles.
  States of strong/weak synaptic weight
  (\ref{fig:multistate_model}) In the multistate model the transition probabilities for potentiation/depression are all equal and it is parameterised by these two values.
  (\ref{fig:binary_model}) The two-state model is parameterised by the two transition probabilities.
  (\ref{fig:cascade_model}) In the cascade model, the transition probabilities decay geometrically with a parameter $x$ (see \cite{Fusi2005cascade}) and synaptic weight takes only two values.
  } \label{fig:models}
\end{figure}

@@

This model is parameterised by the range of values, $q\in[q\lmin,q\lmax]$, for potentiation and depression.
We will use the same values for potentiation and depression in the wild-type as well as potentiation in the \KO\ mutant.
We will use larger values for depression in the \KO\ mutant.
The values of these parameters are listed in \autoref{tab:params}.

\begin{figure}
 \begin{center}
  \aligntop{\includegraphics[height=1.7cm]{pooled.svg}}
 \end{center}
  \caption[The pooled resource model]{The pooled resource model.
  Several two-state synapses share resources that are required for potentiation and depression.
  These resources are depleted as more synapses are potentiated or depressed.
  This pool of synapses can be modelled as one compound synapse.} \label{fig:pooled_model}
\end{figure}

We will describe the effect of training by the decrease in mean synaptic weight:
%
\begin{equation}\label{eq:learning}
  L(t) = \prn{\pr(0)-\pr(t)}\w.
\end{equation}
%
If we imagine something like an LN model for the Purkinje cell, the output would depend on some linear combination of the synaptic weights (weighted by the activities of the corresponding parallel fibres).
If we're not keeping track of synaptic identity, the most natural linear combination to use would be an equal sum of them all.
The behavioural output (VOR gain) will be some unknown, non-linear function of the synaptic weights, so the best we can hope for is to reproduce qualitative features of the experiment, such as whether learning is enhanced or impaired by the mutation or pre-training.

For the most part, we set $f\pot\norm=\half$, $f\pot\inc=f\pot\norm+\Delta f$ and $f\pot\dec=f\pot\norm-\Delta f$, with $\Delta f<0$.
We use the same values for wild-type and mutant and treat gain-increase and decrease symmetrically.
The values of these parameters are listed in \autoref{tab:params}.

We are also assuming that the relation between VOR gain and mean synaptic weight is the same for mutant ant wild-type, except for a linear offset to compensate for the difference in equilibrium weights mentioned above.
This ensures that the qualitative questions (enhancement or impairment of learning) will not be affected.

\section{Simulation and analysis of models}\label{sec:results}

@@

The values for all parameters we will use can be found in \autoref{tab:params}.

@@

\subsection{Step-multistate model}\label{sec:multistate}

@@

\subsection{Two-state model}\label{sec:binary}

@@

\subsection{Linear multistate model}\label{sec:multistate_lin}

In this section, we will consider the multistate model as originally defined in \cite{amit1994learning}, \ie with linearly varying synaptic weight:
%
\begin{equation}\label{eq:multistateLinWeight}
  \w_i = \frac{2i-M-1}{M-1}.
\end{equation}
%
The numerical results can be seen in \autoref{fig:multistate_lin}, but we can get some analytic insight into this model as well.
In essence, this model is like a series of two-state models attached to each other, in contrast to the step-multistate model for which the synaptic weight only changes between one of the pairs of states.

\subsection{Pooled resource model}\label{sec:pooled}

@@

\subsection{Cascade model}\label{sec:cascade}
